---
title: "SMA Epigenomics Pipeline — Analysis Report"
subtitle: "Assessing Genome-Wide Pleiotropic Effects of ASO1+VPA Treatment"
author: "Muna Berhe | MSc Bioinformatics | QMUL"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    fig_width: 10
    fig_height: 7
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(DESeq2)
library(DMRcaller)
library(BSseq)
library(GenomicRanges)
library(rtracklayer)
library(clusterProfiler)
library(org.Hs.eg.db)
library(EnhancedVolcano)
library(pheatmap)
library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(tidyverse)
library(RColorBrewer)
library(DT)

setwd("~/sma_epigenomics_pipeline")

padj_thresh <- 0.05
lfc_thresh  <- 1
```

---

# Project Overview

## Background

Spinal Muscular Atrophy (SMA) is a severe autosomal recessive motor neuron disease
caused by loss-of-function mutations in the *SMN1* gene. Nusinersen (ASO1) corrects
SMN2 exon 7 splicing but also deposits the repressive histone mark H3K9me2 at the
SMN2 locus, partially counteracting its own effect (Marasco et al., 2022).

Valproic acid (VPA), an HDAC inhibitor, counteracts this chromatin compaction.
The ASO1+VPA combination outperforms ASO1 alone — but VPA is a genome-wide
chromatin modifier. **This report characterises those genome-wide pleiotropic effects.**

## Research Aims

| Aim | Analysis | Tool |
|-----|----------|------|
| 1 | Differential gene expression: ASO1+VPA vs ASO1 | DESeq2 |
| 2 | Alternative splicing, including SMN2 exon 7 | rMATS |
| 3 | Differential methylation integrated with expression | DMRcaller |

---

# Differential Expression Analysis

## Load Results
```{r load-deseq2}
res_df <- read.csv("results/differential/deseq2_results.csv", row.names = 1)

n_sig  <- sum(res_df$padj < padj_thresh & abs(res_df$log2FoldChange) > lfc_thresh, na.rm = TRUE)
n_up   <- sum(res_df$padj < padj_thresh & res_df$log2FoldChange > lfc_thresh, na.rm = TRUE)
n_down <- sum(res_df$padj < padj_thresh & res_df$log2FoldChange < -lfc_thresh, na.rm = TRUE)
```

## Summary
```{r deseq2-summary}
cat("Total significant DEGs:", n_sig, "\n")
cat("Upregulated in ASO1+VPA:", n_up, "\n")
cat("Downregulated in ASO1+VPA:", n_down, "\n")
```

> **`r n_sig` genes** were significantly differentially expressed between ASO1+VPA
and ASO1 alone (padj < `r padj_thresh`, |log2FC| > `r lfc_thresh`),
of which **`r n_up` were upregulated** and **`r n_down` were downregulated**
in the ASO1+VPA condition.

## Volcano Plot
```{r volcano, fig.cap="Volcano plot of differential expression. Red points are significant DEGs."}
EnhancedVolcano(res_df,
                lab      = rownames(res_df),
                x        = "log2FoldChange",
                y        = "padj",
                pCutoff  = padj_thresh,
                FCcutoff = lfc_thresh,
                title    = "ASO1+VPA vs ASO1",
                subtitle = "Differential Expression")
```

## MA Plot
```{r ma-plot, fig.cap="MA plot showing mean expression vs log2 fold change."}
counts <- read.table("results/counts/counts.txt", header = TRUE, skip = 1, row.names = 1)
counts <- counts[, 6:ncol(counts)]
coldata <- data.frame(
  condition = factor(c(rep("ASO1", ncol(counts)/2), rep("ASO1_VPA", ncol(counts)/2))),
  row.names = colnames(counts)
)
dds <- DESeqDataSetFromMatrix(countData = counts, colData = coldata, design = ~ condition)
dds <- DESeq(dds)
res <- results(dds, contrast = c("condition", "ASO1_VPA", "ASO1"),
               alpha = padj_thresh, lfcThreshold = lfc_thresh)

plotMA(res, main = "MA Plot — ASO1+VPA vs ASO1", ylim = c(-5, 5), alpha = padj_thresh)
```

## PCA Plot
```{r pca, fig.cap="PCA of RNA-seq samples coloured by condition."}
vsd <- vst(dds, blind = FALSE)
pca_data <- plotPCA(vsd, intgroup = "condition", returnData = TRUE)
percent_var <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(PC1, PC2, color = condition)) +
  geom_point(size = 5) +
  geom_text(aes(label = name), vjust = -1, size = 3) +
  xlab(paste0("PC1: ", percent_var[1], "% variance")) +
  ylab(paste0("PC2: ", percent_var[2], "% variance")) +
  scale_color_manual(values = c("ASO1" = "#0D1B2A", "ASO1_VPA" = "#00897B")) +
  theme_bw() +
  ggtitle("PCA — Sample Clustering by Condition")
```

## Top 50 DEG Heatmap
```{r heatmap, fig.cap="Heatmap of top 50 differentially expressed genes."}
sig_genes <- rownames(res_df)[!is.na(res_df$padj) & res_df$padj < padj_thresh][1:50]
mat <- assay(vsd)[sig_genes, ]
mat <- mat - rowMeans(mat)

annotation_col <- data.frame(
  Condition = coldata$condition,
  row.names = rownames(coldata)
)
ann_colors <- list(Condition = c(ASO1 = "#0D1B2A", ASO1_VPA = "#00897B"))

pheatmap(mat,
         annotation_col    = annotation_col,
         annotation_colors = ann_colors,
         color             = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
         show_rownames     = TRUE,
         show_colnames     = TRUE,
         cluster_rows      = TRUE,
         cluster_cols      = TRUE,
         fontsize_row      = 7,
         main              = "Top 50 DEGs — ASO1+VPA vs ASO1")
```

## Top DEGs Table
```{r deseq2-table}
sig_table <- res_df[!is.na(res_df$padj) & res_df$padj < padj_thresh, ]
sig_table <- sig_table[order(sig_table$padj), ]
sig_table <- round(sig_table, 4)

datatable(sig_table,
          caption = "Significant DEGs (padj < 0.05, |log2FC| > 1)",
          options = list(pageLength = 15, scrollX = TRUE))
```

---

# Differential Methylation Analysis

## Load Results
```{r load-dmrs}
dmrs_df <- read.csv("results/differential/dmrs.csv")
dmrs_df$width <- dmrs_df$end - dmrs_df$start

n_dmrs <- nrow(dmrs_df)
n_gain <- sum(dmrs_df$regionType == "gain", na.rm = TRUE)
n_loss <- sum(dmrs_df$regionType == "loss", na.rm = TRUE)
```

## Summary

> **`r n_dmrs` DMRs** were identified between ASO1+VPA and ASO1,
of which **`r n_gain` gained** and **`r n_loss` lost** methylation
in the ASO1+VPA condition.

## Methylation Difference Distribution
```{r dmr-histogram, fig.cap="Distribution of methylation differences across all DMRs."}
ggplot(dmrs_df, aes(x = methylationDifference)) +
  geom_histogram(bins = 50, fill = "#00897B", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  theme_bw() +
  labs(title    = "Distribution of Methylation Differences",
       subtitle = "ASO1+VPA vs ASO1",
       x        = "Methylation Difference",
       y        = "Count")
```

## DMR Size Distribution
```{r dmr-size, fig.cap="Distribution of DMR widths in base pairs."}
ggplot(dmrs_df, aes(x = width)) +
  geom_histogram(bins = 50, fill = "#0D1B2A", color = "white") +
  theme_bw() +
  labs(title = "DMR Size Distribution",
       x     = "DMR Width (bp)",
       y     = "Count")
```

## DMRs per Chromosome
```{r dmr-chr, fig.cap="Number of DMRs per chromosome."}
chr_counts <- as.data.frame(table(dmrs_df$seqnames))
colnames(chr_counts) <- c("chromosome", "count")

ggplot(chr_counts, aes(x = chromosome, y = count)) +
  geom_bar(stat = "identity", fill = "#00897B") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Number of DMRs per Chromosome",
       x     = "Chromosome",
       y     = "Number of DMRs")
```

## Genomic Annotation of DMRs
```{r dmr-annotation, fig.cap="Genomic distribution of DMRs relative to gene features."}
dmrs_gr <- makeGRangesFromDataFrame(dmrs_df, keep.extra.columns = TRUE)
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
anno <- annotatePeak(dmrs_gr, tssRegion = c(-3000, 3000),
                     TxDb = txdb, annoDb = "org.Hs.eg.db")
plotAnnoPie(anno, main = "Genomic Distribution of DMRs")
```

## Distance to TSS
```{r dmr-tss, fig.cap="Distribution of DMR distances to nearest transcription start site."}
plotDistToTSS(anno, title = "Distance of DMRs to Nearest TSS")
```

## DMR Table
```{r dmr-table}
dmr_table <- dmrs_df[, c("seqnames", "start", "end", "width",
                           "methylationDifference", "regionType", "pValue")]
dmr_table <- dmr_table[order(dmr_table$pValue), ]
dmr_table <- round(dmr_table, 4)

datatable(dmr_table,
          caption = "All DMRs ordered by p-value",
          options = list(pageLength = 15, scrollX = TRUE))
```

---

# DMR-DEG Integration

## Load Overlap Results
```{r load-overlap}
overlap_df <- read.csv("results/differential/dmr_deg_overlap.csv")
n_overlap  <- nrow(overlap_df)
```

## Summary

> **`r n_overlap` genes** have both a nearby DMR and significant differential
expression, suggesting coordinated epigenetic and transcriptional changes
induced by ASO1+VPA treatment.

## Methylation Change vs Expression Change
```{r scatter, fig.cap="Scatter plot of methylation difference vs log2 fold change for overlapping DMR-DEG pairs."}
ggplot(overlap_df, aes(x = methylation_diff, y = log2FoldChange)) +
  geom_point(aes(color = -log10(padj)), size = 2.5, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_gradient(low = "#B2DFDB", high = "#0D1B2A",
                       name = "-log10(padj)") +
  geom_text(data = overlap_df[abs(overlap_df$log2FoldChange) > 2, ],
            aes(label = gene_name), size = 2.5, vjust = -0.5) +
  theme_bw() +
  labs(title    = "DMR Methylation Change vs Gene Expression Change",
       subtitle = "Each point = one DMR-DEG overlap",
       x        = "Methylation Difference (ASO1+VPA - ASO1)",
       y        = "log2 Fold Change (ASO1+VPA / ASO1)")
```

## Overlap Table
```{r overlap-table}
datatable(overlap_df,
          caption = "Genes with both differential methylation and expression",
          options = list(pageLength = 15, scrollX = TRUE))
```

---

# Pathway Enrichment
```{r enrichment, fig.cap="Top enriched GO biological process terms among DEGs."}
deg_genes <- rownames(res_df)[!is.na(res_df$padj) & res_df$padj < padj_thresh]

ego <- enrichGO(gene          = deg_genes,
                OrgDb         = org.Hs.eg.db,
                keyType       = "SYMBOL",
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                readable      = TRUE)

dotplot(ego, showCategory = 20, title = "GO Biological Process Enrichment")
```

---

# Session Information
```{r session-info}
sessionInfo()
```

---

# References

- Marasco et al. (2022) *Cell* 185:2057-2070
- Catoni et al. (2018) *Nucleic Acids Research* 46:e114
- Love et al. (2014) *Genome Biology* 15:550
- Krueger & Andrews (2011) *Bioinformatics* 27:1571-1572